---
title: "Practical Machine Learning Course Project"
author: "Jørgen D. Tyvand"
date: "March 18, 2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

This report covers the final course project for the Practical Machine Learning course from Johns 
Hopkins University on Coursera. The course project concerns data collected from
activity sensors gathered from 6 participants with regards to the correctness of
performing a barbell lift. For further information see the [study website](http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har).

We are to train various machine learning models and report
on accuracy and out of sample error using a cross validation set. 
We also have to test the models on a supplied test set of 20 test cases for a 
separate quiz in the course.

We train 4 models: a classification tree model ("rpart"), a linear discriminate analysis model ("lda"),
a generalized boosted model ("gbm") and a random forest model ("rf"). We find that there is an increase in accuracy
from each of these models to the next (random forest having the highest accuracy), and
that both the boosted and the random forest models correctly predict all 20 test cases 
correctly for the course quiz.

## Data loading and cleaning

The following code loads the required packages and downloads the required files if they
do not exist in the working directory. The files are then read to data frames.
A crucial step here, to ensuring that the training of models does not go on forever, is 
to set all blank values to NA so we can omit columns with blank values later on.

```{r, message = FALSE}
library(caret)
library(dplyr)
if(!file.exists("pml-training.csv")) {
        trainingURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
        download.file(trainingURL, destfile = "pml-training.csv", method = "curl")
}
if(!file.exists("pml-testing.csv")) {
        testingURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
        download.file(testingURL, destfile = "pml-testing.csv", method = "curl")
}
training <- read.csv("pml-training.csv", na.strings = c("NA", ""))
testing <- read.csv("pml-testing.csv", na.strings = c("NA", ""))

print(dim(training))
```

As we can see, the training set contains 19622 rows of 160 observations, one of these being the "classe" variable which we are to predict. This variable is a factor variable of 5 different classes,
given as follows:

- Class A, exercise done exactly according to the specification
- Class B, throwing the elbows to the front
- Class C, lifting the dumbbell only halfway
- Class D, lowering the dumbbell only halfway
- Class E, throwing the hips to the front

For cleaning the data set, we first set the seed for reproducibility, 
and remove any columns with NA values. We also remove the first 7 rows of the data set, 
as these include values like user name and time stamp, which are not really good predictors, as they are not related to the actual performing of the barbell lift.

```{r}
set.seed(2604)
training <- training %>% select_if(~ !any(is.na(.))) %>% select(-c(1:7))
testing <- testing %>% select_if(~ !any(is.na(.))) %>% select(-c(1:7))

print(dim(training))
```

We have now reduced the training set to 53 columns, with 52 predictors and the 
classe variable.

## Data partitioning and model training

To train the models and compute the out of sample error, we split the training set
into a smaller training set (70%) and a cross validation set (30%). 

```{r}
inTrain <- createDataPartition(training$classe, p = 0.7, list = FALSE)
train <- training[inTrain,]
cv <- training[-inTrain,]
```

A lot of previous course participants have had problems with very long run times
when training the models, especially the random forest model. Len Greski (one of the mentors for this course on Coursera) has written
a [report](https://github.com/lgreski/datasciencectacontent/blob/master/markdown/pml-randomForestPerformance.md) on how to speed up the training process, and I have implemented both
the trainControl settings and the parallelization suggested in his report when training
all my models.

```{r, message = FALSE}
library(parallel)
library(doParallel)
cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)
fitControl <- trainControl(method = "cv",
                           number = 5,
                           allowParallel = TRUE)
rpartMod <- train(classe ~ ., data = train, method = "rpart", trControl = fitControl)
ldaMod <- train(classe ~ ., data = train, method = "lda", trControl = fitControl)
gbmMod <- train(classe ~ ., data = train, method = "gbm", trControl = fitControl, verbose = FALSE)
rfMod <- train(classe ~ ., data = train, method = "rf", trControl = fitControl)
stopCluster(cluster)
registerDoSEQ()
```

Printouts of each of the model predictions on the cross validation sets can be found in
the [Appendix]. We see that we get the following accuracy and out of error estimates
for the four models:

- rpart: Accuracy 0.5006, out of sample error 0.4994
- lda:   Accuracy 0.7035, out of sample error 0.2965
- gbm:   Accuracy 0.9669, out of sample error 0.0331
- rf:    Accuracy 0.9932, out of sample error 0.0068

## Prediction on test set

Finally we predict using each model on the test set required for the final course quiz

```{r}
rpartTestPred <- predict(rpartMod, testing)
print(rpartTestPred)
ldaTestPred <- predict(ldaMod, testing)
print(ldaTestPred)
gbmTestPred <- predict(gbmMod, testing)
print(gbmTestPred)
rfTestPred <- predict(rfMod, testing)
print(rfTestPred)
```
I have used the answers for the random forest (highest accuracy) on the course quiz, which 
gives 20/20 correct answers. We therefore see that both the gbm and rf models 
correctly predict the test set.

## Appendix

The following printouts show the confusionMatrix printouts for each of the four models
when predicting the cross validation sets, showing the increase in accuracy up to 99.32% 
for the random forest model.

```{r}
rpartPred <- predict(rpartMod, cv)
print(confusionMatrix(cv$classe, rpartPred))
ldaPred <- predict(ldaMod, cv)
print(confusionMatrix(cv$classe, ldaPred))
gbmPred <- predict(gbmMod, cv)
print(confusionMatrix(cv$classe, gbmPred))
rfPred <- predict(rfMod, cv)
print(confusionMatrix(cv$classe, rfPred))
```